Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-19 MANUSCRIPT.qmd
edits

---
title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
subtitle: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands |  | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
      - Writing – Review & Editing 
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

@@

|              | Variable                  | Source       | Geoprocessing          |
|------------------|-------------------|------------------|------------------|
| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
|              | Mean Temperature\*        | PRISM        | Basin average          |
|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
| Geospatial   | Elevation\*               | DEM          | Basin average          |
|              | Area                      | DEM          | Basin average          |
|              | Slope                     | DEM          | Basin average          |
|              | Aspect                    | DEM          | Percent areal coverage |
|              | Land Cover\*              | NLCD         | Percent areal coverage |
|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
|              | Geology                   | USGS         | Percent areal coverage |
|              | Karst                     | USGS         | Percent areal coverage |

@@

![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

@@

However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act. Even so, many of these gages are not suited for base-flow studies because they emphasize flood monitoring and lack the ability to accurately measure low-flow dynamics. New, non-USGS streamgages are typically installed by flood control districts (e.g. the ALERT system) and are designed for tracking flood flows rather than base flow. New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.

>>>>

---
title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
      - Writing – Review & Editing 
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

@@

|   | Variable | Source | Geoprocessing |
|------------------|-------------------|------------------|------------------|
| Hydroclimate | Precipitation\* | PRISM | Basin average |
|  | Mean Temperature\* | PRISM | Basin average |
|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
| Geospatial | Elevation\* | DEM | Basin average |
|  | Area | DEM | Basin average |
|  | Slope | DEM | Basin average |
|  | Aspect | DEM | Percent areal coverage |
|  | Land Cover\* | NLCD | Percent areal coverage |
|  | Hydrologic Soil Group\* | SSURGO | Percent areal coverage |
|  | Geology | USGS | Percent areal coverage |
|  | Karst | USGS | Percent areal coverage |

@@

![SHAP value plot of features used in final model. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

@@

However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus. There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act. Even so, many of these gages are not suited for base-flow studies because they emphasize peak flow monitoring and lack the ability to accurately measure low-flow dynamics [@flood-control2020]. New, non-USGS streamgages are typically installed by flood control districts (e.g. the ALERT system) and are designed for tracking flood flows rather than base flow. New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.
